Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-03-27 paper/tidy-missing-data-paper.Rmd
condence points

1. **Missing value labels**. A simple label to the shadow matrix clear, "NA" and "!NA" for missing and not missing, represented as 0 and 1.

The $B$ matrix is a very common representation of missing values. It can be created in the "R" language with \code{is.na(data)}. The values 0 and 1 (or FALSE and TRUE) can be confusing, however, as it is not immediately clear what they represent: does 0 indicate a the absence (0) of observation, or the presence (1) of a missing value? This is solved by adding a simple label to the shadow matrix: "NA" and "!NA" for missing and not missing. Values in the shadow matrix can be represented as "factor" types values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing (Figure \@ref(fig:shadow-matrix-progress) parts 1-3).

1. Special missing values: Building on 1, the values in the shadow matrix can be "special" missing values, indicated as `NA_instrument`, where "instrument" is a very short label indicating an instrument error resulting in a missing values.

1. Coordinated names: Variables in the shadow matrix gain the same name as in the data, with the suffix "_NA".

1. Cohesiveness: Binding the shadow matrix column-wise to the original data creates a cohesive _nabular_ data format, so that it cannot be out of sync with the data. useful for visualization, summaries, and tracking imputed values.

(**note - change R matrix to something else**)

### Missing value labels {#missing-labels}

### Special missings {#special-missings}

()

Extending missing value labels to include "special" missing values labelled as "NA_\<suffix\>" allows for representing additional missingness features such as instrument failure and drop-out as "NA_instr_fail", and "NA_drop_out", and even imputations (Table \@ref(tab:shadow-encoding)). The underlying number representation changes from binary to integer, allowing for unlimited missingness features. For example, instrument failure and drop out would be represented as 2 and 3 (Table \@ref(tab:shadow-encoding)). Encoding special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in \pkg{naniar} (\@ref(verbs-recode)).

@@

@@

## Coordinated names {#coordinated-names}

A consistent short suffix "`_NA`" for each variable keeps names coordinated throughout analysis (Figure \@ref(fig:shadow-matrix-progress) part 4). The suffix is short and easy to remember in analysis or visualization. It also makes a clear distinction with `var_NA` being a random variable of the missingness of a variable, `var`. This subtle change is important as it shifts the focus from the value of a variable to its missingness state, making intent clear when performing analysis.

(make this it's own section)

>>>>

1. **Missing value labels**. Simple labels, "NA" and "!NA", for missing and not missing, clearly identifies missing values for analysis and plotting. This is an improvement over the commonly used binary matrix ($B$), which makes it unclear if 0 indicates the absence (0) of observation, or the presence (1) of a missing value. Missing value labels can be represented as "factor" types values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing (Figure \@ref(fig:shadow-matrix-progress) parts 1-3).

1. **Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA_\<suffix\>" - for example: `NA_instrument` uses a short label, "instrument", indicating instrument error resulting in missing values. These could be also used to indicate imputations, (Table \@ref(tab:shadow-encoding)). The underlying number representation changes from binary to integer, allowing for unlimited missingness features. For example, instrument failure and drop out would be represented as 2 and 3 (Table \@ref(tab:shadow-encoding)). Encoding special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in \pkg{naniar} (\@ref(verbs-recode)).

@@

1. **Coordinated names**: Variables in the shadow matrix gain a consistent short suffix, "_NA", keeping names coordinated throughout analysis (Figure \@ref(fig:shadow-matrix-progress) part 4). The suffix is short and easy to remember in analysis or visualization. This shifts the focus from the value of a variable to its missingness state, making intent clear when performing analysis.

(It makes a clear distinction with `var_NA` being a random variable of the missingness of a variable, `var`. )

1. **Cohesiveness**: Binding the shadow matrix column-wise to the original data creates a cohesive _nabular_ data format, so that it cannot be out of sync with the data. useful for visualization, summaries, and tracking imputed values. This is discussed in more detail below in \@ref(nabular-data).

@@
